Combining global tree cover loss data with historical national forest-cover maps to look at six decades of deforestation and forest fragmentation in Madagascar

### 2017-05-11 deforestmap.Rmd
Improve discussion

### 4.1 Evolution of forest-cover in Madagascar over the last 60 years

There has been a long debate to know whether Madagascar had lost 90% of its forest since the arrival of the humans on the island [@Virah-Sawmy2009, @Agarwal2005]. What it sure now, is that Madagascar has lost around 50% of its natural forest in six decades since 1953.

@@

The three previous studies have used (i) different forest definition (associated to independant visual interpretations of observation polygons to train the classifier), (ii) different classification algorithms (Maximum Likelihood Classification for @Harper2007 and @MEFT2009 vs. Random Forests for @ONE2015), and (iii) do not systematically correct for the presence of clouds. Despite these differences, the four studies gave similar forest-cover estimates and show the same deforestation trend over Madagascar, with a decrease of the deforestation rate on the period 1990-2005 followed by a progressive increase of the deforestation rate after 2005. Given the fact that our deforestation rate estimates have been computed from cloud-free maps, our estimates can be considered more accurate compared to estimates from previous studies, in particular for deforestation rates expressed in number of hectares per year. Current rates of deforestation for Madagascar (98651 ha/yr, 1.08 %/yr, if we consider the period 2010-2014) could be used to build reference scenarios for deforestation in Madagascar [@Olander2008]. Forest-cover and deforestation rate estimates from this study could also be used to update the estimates in the upcoming FAO Forest Resources Assessment reports [@Keenan2015].

### 4.2 Consequences of deforestation and forest fragmentation on biodiversity and carbon emissions

@@

### 4.3 Interest of such an approach and limitations

- simple
- easilly reproducible
- no need to define a tree cover threshold to

>>>>

### 4.1 Interest of combining recent global annual tree cover loss data with historical national forest-cover maps

In this study, we combined recent (2000-2014) global annual tree cover loss data [@Hansen2013] with historical (1953-2000) national forest-cover maps [@Harper2007] to look at six decades (1953-2014) of deforestation and forest fragmentation in Madagascar. With this approach, we obtained annual cloud-free 30m-resolution forest-cover maps for Madagascar for the period 2000 to 2014. This approach overcome the problem associated to the forest definition which is necessary to transform the 2000 global tree cover data from @Hansen2013 into a forest/non-forest map. Instead, an historical national or regional forest-cover map, based on a more appropriate forest definition at the national or regional scale, can be used as a mask to derive future forest-cover maps until 2014 taking into account global annual tree cover loss data by @Hansen2013. This approach could be easilly extended to other regions or countries with available forest-cover maps at one date on the period 2000-2014. Moreover, that approach can be repeated in the future to obtain future forest-cover change maps, as and when the @Hansen2013 product is updated. The R/GRASS code used for this study is made freely available in a GitHub repository (see Data availability statement). Code should be easilly transposable to other study sites or re-used for future analysis in Madagascar. Of course, the accuracy of the derived forest-cover change maps depends directly on the accuracy of the historical forest-cover maps and the accuracy of the tree cover loss data. Accuracy of the tree cover loss data [@Hansen2013] has been discussed in several recent studies [@Verhegghen2016].

@@

### 4.2 Evolution of forest-cover in Madagascar since 1953 and current forest-cover

We estimated that natural forests covered 8.9 Mha in Madagascar (15% of the national territory of 587041 km<sup>2</sup>) in 2014 and that Madagascar has lost 44% of its natural forest in six decades since 1953 (37% since 1973). There is an age-old debate about the original forest cover in Madagascar to know to which extent humans have altered the natural forest landscape since their large-scale settlement around 800 CE [@Cox2012]. French naturalists Henri Humbert [@Humbert1927] and Henri Perrier de la Bâthie [@Perrier1921] stated that the full island was originally covered by forest, leading to the common statement that 90% of the natural forests have disappeared since the arrival of humans on the island [@Kull2000]. More recent studies counter-balanced that point of view saying that grasslands existed in Madagascar long before human arrival and were determined by climate, natural grazing and other natural factors [@Virah-Sawmy2009, @Vorontsova2017]. Other authors have questioned the entire narrative of extensive alteration of the landscape by early human activity which, through legislation, has severe consequences on local people [@Klein2002; @Kull2000]. Whatever the original proportion of natural forests and grasslands in Madagascar, our results demonstrate that human activities since the 5Os have profoundly impacted the forest landscape and that recent conservation and development programs in Madagascar have failed to stop deforestation in the recent years. Associated to deforestation, we showed in our study that residual forest in Madagascar were highly fragmented with 95% of the forest being at a distance <1.2 km of the forest edge in 2014. Both deforestation and forest fragmentation has strong consequences on biodiversity and carbon emissions in Madagascar. Around 90% of Madagascar's endemic species are forest dependent [@Allnutt2008; @Goodman2005] and @Allnutt2008 estimated that deforestation between 1953 and 2000 had led to an extinction of 9% of the species.

Increased wind speed, vapour pressure deficit and disturbance pressure in the vicinity of forest edges increase tree mortality, resulting in persistent carbon emissions long after deforestation occurs. Fire, wind and desiccation can penetrate up to kilometres into the forest15,19,20, altering microclimate within 100-300m of the forest edge13,21,22,23. Field studies have estimated relative carbon losses in forest edges (e) averaging 11% (ref. 24) and up to 36% (ref. 22) and 50% (ref. 25) of aboveground biomass within the 100m edge zone, and modelling studies have shown biomass reductions of up to 70% in small fragments2

@@

More efforts need to be done to save the few remaining natural tropical forests.

### 4.3 Deforestation trend and causes of the acceleration of the deforestation since 2005

In our study, we have shown that the progressive decrease of the deforestation rate on the period 1990-2005 was followed by a continuous increase in the deforestation rate on the period 2005-2014. We showed that deforestation rate has more than doubled on the period 2010-2014 (98651 ha/yr, 1.08%/yr) compared to 2000-2005 (42296 ha.yr, 0.43%/yr). These results are confirmed by previous studies on deforestation at the national scale for Madagascar [@Harper2007; @MEFT2009; @ONE2015]. Despite differences in the methodologies regarding (i) forest definition (associated to independant visual interpretations of observation polygons to train the classifier), (ii) classification algorithms (Maximum Likelihood Classification for @Harper2007 and @MEFT2009 vs. Random Forests for @ONE2015), (iii) deforestation rate computation, and (iii) correction for the presence of clouds, these studies gave similar forest-cover estimates and show the same deforestation trend over Madagascar. The increase of the deforestation rates after 2005 could be attributed to both population growth and political instability in the country. Nearly 90% of Madagascar's population relies on biomass for their daily energy needs [@Minten2013] and the link between population size and deforestation has previously been demonstrated in Madagascar [@Vieilledent2013, @Gorenflo2011]. With a mean demographic growth rate of about 2.8%/yr and a population which has increased from 16 to 24 million people on the period 2000-2015 [UN2015], the increasing demand in wood-fuel and space for agriculture is likely to explain the increase in the deforestation rate. The political crisis of 2009, followed by several years of political instability and weak governance is also very likely to explain the increase in the deforestation rate observed on the period 2005-2014. Our deforestation rate estimates from 1990 to 2014 have been computed from cloud-free maps and can be considered more accurate in comparison with estimates from the previous studies, in particular for deforestation rates expressed in number of hectares per year. Forest-cover and deforestation rate estimates from this study could be used to update the estimates in the upcoming FAO Forest Resources Assessment reports [@Keenan2015]. Current rates of deforestation for Madagascar (98651 ha/yr, 1.08 %/yr, if we consider the period 2010-2014) could be used to build reference scenarios for deforestation in Madagascar [@Olander2008].
